Declarative Pearl: Lightweight Probabilistic Programming in Curry

### 2015-08-18 Paper/paper.tex
Start with quering a bayesian network example

In the style of \cite{pfp}, the library's heart is a data type that
represents distributions: such a distribution consists of pair of
value and its probability. %

\begin{verbatim}
data Probability = Prop Float
data Dist a = Dist (a,Probability)
\end{verbatim}

However, the attentive reader may notice a slight difference: the PFLP
approach does not need to represent the collection of possible values
and their corresponding probabilities as list. %
In functional languages like Curry, the representation of several
results is commonly represented with nondeterminism instead of
directly using a container like lists. %
That is, we model probabilistic values as nondeterministic pairs of
possible value tagged with their probability. %

For example, we can define a biased coin with a $30\%-70\%$ chance to
land on head and tail, respectively, as follows. %

\begin{verbatim}
data Coin = Head | Tail
biasedCoin = (Head, 0.3)
biasedCoin = (Tail, 0.7)
\end{verbatim}

@@

\begin{verbatim}
fairCoin = (Head ? Tail,0.5)
\end{verbatim}

\begin{itemize}
\item Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  Based on which criteria do you choose the host language? %
  As for probabilistic languages that started growing over time,
  libraries that enable probablistic programming exist in the whole
  variety of paradigms: declarative as well as imperative. %
  For a start, you obviously choose the paradigm that has the clearest
  structure and enables straight-forward programming. %
  We want to address a wide range of developers and even
  non-PL-experts, thus, we choose a paradigm than can be learnt
  easily. %
  Thus, obvioulsy, you choose a declarative programming language. %
  By choosing Curry, we argue that it is sufficient to reuse built-in
  features in order to cover the essence of probabilistic programs,
  especially its semantic. %

@@

>>>>

In the style of \cite{pfp}, the library's heart is a data type that
represents distributions: such a distribution consists of a value and
its probability. %

\begin{verbatim}
data Probability = Prop Float
data Dist a = Dist a Probability

mkDist :: a -> Probability -> Dist a
mkDist = Dist
\end{verbatim}

However, the attentive reader may notice a slight difference: the PFLP
approach does not need to represent a distribution as a list of its
possible values. %
In functional logic languages like Curry, the representation of
several results is commonly represented with nondeterminism instead of
directly using a container like lists. %
That is, we model probabilistic values as nondeterministic pairs of
possible value tagged with their probability. %

For example, we can define a biased coin with a $30\%$ vs $70\%$
chance to land on head and tail, respectively, as follows. %

\begin{verbatim}
type Coin = Bool

head, tail :: Coin
head = True
tail = False

biasedCoin = (head, 0.3)
biasedCoin = (tail, 0.7)
\end{verbatim}

@@

\begin{verbatim}
fairCoin = (head ? tail,0.5)
\end{verbatim}

Let us now take a look at the bayesian network example again. %
We have already seen the definition of the model in the
introduction. %
However, the more interesting part is to make queries on the given
model that cannot be answered with the probability table itself. %
For example, we can ask the model for the probability that a burglary
occured given the fact that the alarm was set of. %

\begin{verbatim}
-- P(B=T | A=T)
alarmByBurglary =
  burglary >>>= \b ->
  earthquake >>>= \e ->
  alarm b e >>>= \a ->
  guard a >>>= \_ ->
  pure b
\end{verbatim}

@@

\item Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  Based on which criteria do you choose the host language? %
  As for probabilistic languages that started growing over time,
  libraries that enable probablistic programming exist in the whole
  variety of paradigms: declarative as well as imperative. %
  For a start, you obviously choose the paradigm that has the clearest
  structure and enables straight-forward programming. %
  We want to address a wide range of developers and even
  non-PL-experts, thus, we choose a paradigm than can be learnt
  easily. %
  Thus, obvioulsy, you choose a declarative programming language. %
  By choosing Curry, we argue that it is sufficient to reuse built-in
  features in order to cover the essence of probabilistic programs,
  especially its semantic. %
